Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-06-29 POPL/paper.tex
POPL Paper Stub

% -*- fill-column: 80; -*-
\documentclass[sigconf,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{stmaryrd}

\usepackage{todonotes}
\newcommand{\todoin}[1]{\todo[inline]{TODO:\@ #1}}

\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\nat}{\mathbb{N}}

@@

\newif\ifdraft
\drafttrue
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven}: #1]}\fi}
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro}: #1]}\fi}
\newcommand{\pipe}{\,|\,}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

@@

\title{Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
\author{Steven Schaefer}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{stschaef@umich.edu}

@@

\begin{abstract}
  We propose a semantic framework for the study of formal grammars. First, we
  provide a syntax-independent notion of formal grammar as a function from
  strings to sets of parse trees, generalizing the familiar notion of a language
  as a set of strings. This point-of-view naturally endows a grammar with the
  structure of a presheaf, which allows many language-theoretic constructs to be
  defined via universal properties.

  With these categorical constructions in mind, we propose a new syntactic
  formalism for formal grammars: a version of dependent linear-non-linear type
  theory where the tensor product is non-commutative. The linear types in this
  theory are interpreted as grammars --- and linear terms as parse transformers.
  The non-dependent fragment of this type theory is already expressive enough to
  capture regular expressions, context-free expressions, and finite automata.
  Moreover, the isomorphism between regular languages and finite automata is
  definable as a proof term in our formalism. The
  non-dependent fragment of this type theory is expressive enough to capture
  both regular and context-free expressions, as well as finite automata already
  regular expressions, context-free expressions, and finite automata. Moreover,
  the is definable as a proof term in this formalism. Further, through use of
  dependency of linear types on non-linear types, we can express richer grammar
  formalisms such as indexed grammars, pushdown automata, and Turing machines.

  Classically, many results in both language and automata theory are proved
  informally with respect to some ambient foundation --- e.g. Kleene algebra, grammars,
  parsers, semantic actions, etc. The type theory provided in
  this paper serves as an intuitive and general framework to syntactically unify
  these related ideas and internalize their
  theorems formally. We give a new form of \emph{logical characterization} of
  grammar classes based on a substructural logic rather than ordinary
  first-order logic. Further, the type theory can not only
  express the grammars as types, but equivalences between grammars and
  parsers as terms of the grammar.

  We give this type theory a semantics in the category of grammars and
  prove a related canonicity theorem. Moreover, we keep an eye toward the
  implementation of this type theory for correct-by-construction parsers and
  parser combinators.
\end{abstract}

@@

\section{A Syntax-Independent Notion of Syntax}
\todoin{Rework intro from LICS draft with a focus on implementation}

\section{Kleene Categories}
\todoin{Tell a story winding through the use of KAs, action algebras, etc to
  motivate why we'd want categorical semantics for KCats}
\todoin{Incoporate Kleene Category definition}
\todoin{End with the punch that $\Set^{\Sigma^{*}}$ is a KCat}

\section{Automata}

\end{document}

>>>>

% -*- fill-column: 80; -*-
\documentclass[sigconf,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{stmaryrd}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows, fit}

\usepackage{pdfpages}

\usepackage{todonotes}

\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\Sigma^{*}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\nat}{\mathbb{N}}

@@

\newif\ifdraft
\drafttrue
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max says}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro says}: #1]}\fi}
\newcommand{\pipe}{\,|\,}

\newcommand{\todoin}[1]{\ifdraft{\todo[inline]{TODO:\@ #1}}\fi}

@@

\pagestyle{plain}

\pagebreak

\ifdraft{
\listoftodos{list of todos}
\pagebreak}\fi

\title{Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}

\author{Steven Schaefer}
\affiliation{\department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{stschaef@umich.edu}

@@

\makeatletter
\let\@authorsaddresses\@empty
\makeatother

\begin{abstract}
  In this paper, we propose a new syntactic formalism for formal grammars: a
  dependent linear/non-linear type theory in which the tensor product is
  non-commutative. The linear types in this calculus may be interpreted as
  grammars, and linear terms as parse transformers. This provides a logical
  characterization of grammar classes based on a substructural logic rather than
  ordinary first-order logic.

  Internally to our calculus, we carry out many classical constructions from
  the theory of formal grammars. Inspired by the common
  structure underlying various ideas in formal language
  theory, our type theory makes these relationships explicit
  and provides a unifying syntax to reason about formal
  grammars.

  We give this type theory a semantics in the category of grammars and prove a
  canonicity theorem showing that every term in a context corresponding to a
  fixed string is equal to a term in a canonical form encoding a parse tree of
  that string. Finally, we discuss the application to the design and
  implementation of correct-by-contruction parsers.
\end{abstract}

@@

\setcounter{page}{1}

\section{Introduction}
The notion of formal language is central to the theory of parsing. A
\emph{formal language} $L$ over an alphabet $\Sigma$ is classically defined as a
\emph{subset} of strings $L \subseteq \String$. This definition is
especially useful as it gives a semantics to formal grammars that is completely
independent of any particular syntactic grammar formalism. Any new
notion of grammar can be given a language semantics and it provides a precise
mathematical specification for implementing a \emph{recognizer} of a language.

That is, a recognizer should be extensionally equivalent to the indicator function
$\chi_{L} : \String \to \Prop$ mapping each string $w$ to the the proposition
that it is in the language $w \in L$. However, language recognition is
insufficient for specifying a parser --- a function that decomposes a string
into semantic components that adhere to the structure of a formal grammar. In
practice, parsers do more than simply return whether a string is valid or
invalid; instead, they emit structured \emph{parse trees}\footnote{Such parse
  trees are usually not materialized in memory, as semantic actions often perform
  computation over the trees while they are being constructed.}

% To aid in the development of verified parsers, the primary contribution
% of this paper is
% an type
% theory whose types natively capture the semantics of formal
% grammars. Types in this theory are grammars, and terms are
% parse transformers. This kind of type theory naturally
% requires (non-commutative) linear and dependent types.
% Within this formalism, we may define parsers as terms
% defined in the context of an input string. Moreover, by virtue of
% typechecking, these parsers are \emph{correct-by-construction}. The intrinsic
% verification gained from the type system therefore removes the need for a
% post-hoc verification in a language like Coq, thus lowering the barrier to
% compiler frontend verification.

% Substructural logics, like linear and separation logic, offer precise control
% over how values from a type may be used, in particular how many times a value
% may be used in a derivation. This control enables \emph{resource-sensitive}
% reasoning, where we may treat a resource as \emph{consumed} after usage. In the
% context of parsing, we may treat the characters composing a string as finite resources
% that are consumed at parse-time.

% Dependent type theory extends the simply-typed $\lambda$-calculus by allowing type
% definitions to contain the occurrence of a term. In lieu of set-based
% foundations, this extension can then be taken as a foundation for mathematics.
% The expressive power unlocked by this insight has enabled the development of
% dependently-typed proof assistants
% such as Coq or Agda. Within each of these languages, systems may be given very
% precise functional specifications of their behavior.

% In 1995, Benton et al.
% \cite{bentonMixedLinearNonlinear1995} first gave categorical models of a logic
% that synthesized linear and non-linear logic. Later, Krishnaswami et al.
% \cite{krishnaswami_integrating_2015} built upon this work to develop a type
% system that unifies both linear and
% dependent types --- referred to as a \emph{linear-non-linear} (LNL) type theory. In
% such settings, we have two sub-theories --- one linear and one non-linear. The
% linear sub-theory embeds ordinary linear types without dependence, while the
% non-linear sub-theory embeds standard Martin-L\"of type theory. The interplay of
% linear types with dependence lives at the boundary between these two fragments.
% Benton et al.\ gives an accounting of how we may jump between these
% fragments via a monoidal adjunction $F \dashv G$ between linear terms and
% non-linear terms.

% To the end of defining real verified
% parsers, in our LNL theory we define a parser as a term,

% \crefname{equation}{}{}
% \begin{equation}
%   \label{eq:parser}
%   w : \String \vdash \mathsf{parse}_{g}(w) : g \oplus \top
% \end{equation}

% We will investigate the precise meaning of judgments like \cref{eq:parser} in
% \cref{sec:tt}, but first let's get an intuitive feel for the meaning behind such a
% term. On the left of the turnstile occurs a single string $w$, we may think of
% this describing what resources are linearly available to us. Using each piece of
% our the context on the left of the turnstile, we build a term
% $\mathsf{parse}_{g}(w)$ of type $g \oplus \top$. Notably, the type being
% instantiated is a sum type, and the data of which side we inject into denotes
% whether the string is accepted or rejected by the parser. That is all to say, a
% parser takes in an input string and either matches it to the desired grammar or
% rejects the string.

% Practical parsers are often implemented via an abstract automaton. For instance,
% a parser may translate a grammar into a state machine recognizes the
% same language as the original grammar, as the automata
% point-of-view lends itself to straightforward execution. The
% strength of our syntax is that it allows us to describe
% these automata as grammars themselves. Moreover, the
% equivalence between the grammar and automata presentations
% is derivable as a proof term within the theory. This allows
% us access to simultaneously internalization of the equivalence
% between grammars and automata as well as the ability to
% export to automaton implementations of a parser that are executable.

% To realize an executable implementation of our type theory,
% we shallowly embed it in Agda. Because our implementation is
% defined as a function in Agda, we can leverage the Agda
% ecosystem for code generation. Further, in lieu of
% implementing a complete language from scratch, we make use of the
% Agda proof system as a setting to guarantee correctness of
% our generate parsers.

% Within this paper, we give our progress in defining a verified
% parser for regular expressions. The construction of such a parser entails
% the definition of a parser for deterministic finite automata (DFA) grammars,
% and an equivalence proof between DFAs and regular expressions. We have thus far
% completed the construction of a parser for DFAs $D$,

% \[
%   w : \String \vdash \mathsf{parse}_{D}(w) : D \oplus \top
% \]

% and half of the equivalence proof between DFAs and regular expressions, both
% formalized in Agda. Upon completion of the second half of this proof, we would
% have a system where a user could simply give a regular grammar and then extract
% an intrinsically verified parser for it. In future work, we
% hope to show an analogous construction involving pushdown
% automata (PDAs) and context-free grammars.

% The contributions of this paper include an LNL theory for reasoning about formal
% grammars, the encoding of automata as formal grammars in this theory, and the
% definition of parser terms for DFA grammars. The contents include,

% \begin{itemize}
%   \item A syntax-independent definition of formal grammar
%   \item The formal rules governing our type system
%   \item Interpretations of the type system for reasoning about formal grammars
%         and formal languages
%   \item The presentation of automata as grammars
%   \item The definition of a parser within our type system
% \end{itemize}

% \section{A Syntax-Independent Notion of Syntax}
% \label{sec:synindsyn}

% % \max{this section doesn't flow naturally from the intro, which is all about the
% %   type theory and parsing. This is where you want to outline in the intro and
% %   possibly say here that we want to revisit ``formal grammar theory'' and then
% %   we will make a type theory that has formal grammar theory as its semantics.}

% In this section we will revisit some foundational ideas in formal
% grammar theory, before giving a type
% theory for which formal grammars serves as its semantics.

% Formal languages and grammars are among some of the oldest and most studied areas in
% theoretical computer science. The developments of the 1950s and 1960s gave rise
% to expressive formalism for reasoning about grammars, and further practical
% algorithms for constructing parsers for regular and context-free grammars.

% A grammar $G$ is often described generatively as a set of production rules for
% generating strings. Often this is presented as a finite set of nonterminal
% symbols $N$, a finite set of terminal symbols $\Sigma$, and a finite set of rewriting
% rules for substitution of nonterminals. In this setting, we continually apply
% rewrite rules until we are left only with a string of terminals
% $w \in \String$. We will give a more precise notion of
% grammar in a moment, but for now let us take this intuitive view of grammars as
% string generators.

% One of the primary objects of study is that of a \emph{formal language} over
% a fixed alphabet $\Sigma$. A language $L$ is a \emph{subset} of strings
% $L \subseteq \String$. This notion of formal language gives an intuitive
% set-based semantics for grammars. That is, we may interpret a grammar $G$ as the
% language $L(G)$ that is generated by it. We are yet to give any precise
% definition of what we mean by ``grammar'', although we note that any suitable
% notion of grammar can be given a language semantics as above.

% Formal languages serve as a basis for mathematically specifying a
% \emph{recognizer} --- a
% function $r : \String \to \Bool$ that determines if a string belongs to a
% given language. However, they are insufficient for specifying a \emph{parser}
% --- a function that decomposes a string into semantic components that adhere to
% the structure of a formal grammar. In practice, parsers don't just tell you
% if a string is valid or invalid. Instead, they emit structured data such as
% abstract syntax trees (ASTs).

% This motivates our first conceptual contribution, a simple and
% syntax-independent notion of formal grammar:

% \begin{definition}
%   \label{def:grammar}
%   A \emph{formal grammar} $G$ over a fixed alphabet $\Sigma$ is a function
%   $G : \String \to \Set$.
% \end{definition}

% We say that a grammar $G$ associates to every string $w$ the set\footnote{We
%   make little comment as to which foundations we use to encode sets. We argue\max{drop ``We argue''}
%   that our construction is agnostic to the particular foundation, and so without
%   loss of generality\max{incorrect use of ``without loss of generality''} we may take a class of small sets, a Grothendieck universe,
%   a type theoretic universe, or any similar foundation. Here we choose a type
%   theoretic universe.} $G w$ of parse
% trees showing that $w$ matches $G$.

% The simplest class of formal grammar on the Chomsky\max{first mention of Chomsky hierarchy, why not bring it up earlier when you are describing what a grammar is? Also cite}
% hierarchy are the \emph{regular expressions}

% \max{inaccurate reflection of common terminology. These are regular expressions,
%   regular grammars usually means a CFG with a restriction on recursion that
%   ensures you only allow Kleene star.}.

% \begin{gather*}
%   \mathsf{Char} \in \Sigma \\
%   \mathsf{Reg} ::= \mathsf{Char}~|~I~|~\mathsf{Reg} \otimes \mathsf{Reg}~|~\mathsf{Reg} \oplus \mathsf{Reg}~|~\mathsf{Reg}^{*}
% \end{gather*}

% Here, we use a non-standard syntax inspired by linear logic. Disjunction of
% regular expressions is written $\oplus$, and concatenation of regular
% expressions is written like the linear tensor product $\otimes$. The
% empty grammar is sometimes written in the same manner as the empty string,
% $\varepsilon$, but we instead choose to represent the empty grammar by $I$. This
% is also linear logic syntax to represent the unit $I$ of the operation $\otimes$.

% To see an illustration of the notion of grammar given in \cref{def:grammar}, consider the regular expression
% $G = (a \otimes b)^{*} \otimes a$ --- a sequence of 0 or more $ab$'s followed by
% an $a$. We may think of $G$ as a function that takes in strings and outputs all
% of the parse trees of that string. Passing the input string $baa$ through $G$
% would output the empty set, because there are no $G$-parses of $baa$. However,
% passing $aba$ through $G$ would output a singleton set, as there is only a
% single $G$-parse of $aba$. Namely, the one that breaks up $aba$ as $(ab)a$.

% Observe that for any input string $w$ for the above grammar $G$, $G(w)$ is
% either empty or a singleton set. Either the string isn't parsed by $G$,
% or it is parsed by $G$ for exactly one reason. That is, $G$ is
% \emph{unambiguous}. If we instead consider the grammar
% $H = (a \otimes b)^{*} \oplus (b \otimes a)^{*}$ there are now ambiguities. For
% instance, what is the action of $H$ on the empty string $\varepsilon$?
% The empty string $\varepsilon$ matches both $(a \otimes b)^{*}$ and $(b \otimes a)^{*}$, so
% $\varepsilon$ can match either the left or the right pattern
% of $H$. As a result, $H(\varepsilon)$ is a two-element set.

% For every grammar, we can specify a formal language recognized by that grammar. By ranging over all input strings, we can
% see that every formal grammar $G$ induces a formal language $L_{G}$,

% \[
%   L_{G} = \{ w \in \String : G w \text{ is inhabited} \}
% \]

% As they are just sets, formal languages naturally are
% endowed with a partial
% order via the subset inclusion ordering. Similarly, formal grammars are
% naturally endowed with the structure of a \emph{category} where a morphism
% $f : G \to H$ is given by a \emph{family} of functions:

% \[
%   f^{w} : Gw \to Hw, w \in \String
% \]

% Intuitively, we read a morphism of grammars as a \emph{parse transformer}: $f$
% translates a
% $G$-parse of $w$ into an $H$-parse of $w$. The notion of isomorphism induced by
% this categorical structure encodes that two grammars are equivalent if there is
% a bijective translation of parses. Isomorphisms in the category of
% grammars are precisely Chomsky's notion of \emph{strong equivalence} of
% grammars \cite{chom1963}.

% The category of grammars is quite rich in structure. It can
% equivalently be described as the functor category $\Set^{\String}$, where
% $\String$ is a discrete category. That is, the objects in $\String$ are strings
% and the only morphisms are identity morphisms. Such functor categories into
% $\Set$ --- often called presheaf categories --- carry a remarkable amount of
% structure. In fact, any presheaf category naturally models dependent type
% theory, which we exploit to motivate the design of the formalism
% presented in \cref{sec:tt}.

% \section{A Type Theory to Reason About Formal Grammars}
% \label{sec:tt}

% \subsection{Syntax}
% \label{subsec:syntax}

% In this section we will walk through most of the syntax of our type theory. We
% include the full syntax in \cref{sec:infer}.

% The linear typing judgment in our syntax takes on the following schematic form,
% \[
%   \Gamma ; \Delta \vdash M : A
% \]

% Let us briefly digest what this judgment is trying to say. First, $A$ represents
% a \emph{linear type} in our syntax. The intended semantics of these linear types
% are formal grammars. That is, the linear typing system is designed to syntactically
% reflect the behavior of formal grammars. For this reason, we may often
% interchangeably use the terms ``linear type'' and ``grammar''. To see the
% correspondence between types and grammars, note that the constructors for linear types
% mirror operations on formal grammars. For example, we have linear type constructors
% $\otimes$, $\oplus$, and $\amp$ which correspond to grammar concatenation,
% disjunction, and conjunction, respectively. The other type constructors also have
% corresponding grammar intuition which may be seen in \cref{subsubsec:other}.

% $M$ is an inhabitant of type $A$, which we want to think of as a parse tree of
% the grammar $A$. The core idea of this entire paper follows precisely from this
% single correspondence: grammars are types, and the inhabitants of these types
% are parse trees for the grammars!

% $\Gamma$ represents a \emph{non-linear context} --- also referred to as an
% \emph{intuitionistic} context. The non-linear fragment of the theory
% corresponds to usual Martin-L\"of type theory. That is, the intuitionistic
% contexts obey all of the usual structural rules --- contraction, weakening, and
% exchange --- so we may use their contents as many times as we'd like.

% On the other hand $\Delta$ represents a \emph{linear context}. These linear
% contexts
% behave substructurally --- in particular they have no contraction, weakening, or exchange
% rule. This means that everything within $\Delta$ must be used \emph{exactly
%   once} and \emph{in order of occurrence}. Thus, we can think of the linear
% contexts as an ordered list of limited
% resources. Once a resource is consumed, we cannot make reference to it again.
% We may then
% think of the variables in a linear context akin to building blocks for
% constructing patterns over strings.

% In the realm of parsing, our substructural
% restrictions make it so that the linear contexts behave quite similarly to
% formal grammars. For instance, linear context extension mirrors concatenation of
% grammars which may be seen via the introduction rule for $\otimes$,

% \[
%   \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta , \Delta' \vdash e \otimes e' : A \otimes B}
% \]

% Just as two grammars may be concatenated, if we can build an $A$ with the
% contents of $\Delta$, and a $B$ with the contents of $\Delta'$, then we we may
% concatenate $\Delta$ and $\Delta'$ to build an inhabitant $A \otimes B$.

% This is perhaps better illustrated through a concrete example. Consider the
% regular expression $A = (a \oplus b) \otimes (b \oplus c)^{*}$ and the string
% $w = abc$. We then have the following derivation that $w$ matches $A$,

% \begin{equation}
%   \label{eq:samplederiv}
%   \inferrule
%   {
%     w_{1} : a \vdash \mathsf{inl}(w_{1}) : a \oplus b \\
%     \inferrule
%     {
%       w_{2} : b \vdash \mathsf{inl}(w_{2}) : b \oplus c \\
%       w_{3} : c \vdash \mathsf{inr}(w_{3}) : b \oplus c
%     }
%     {
%       w_{2} : b, w_{3} :c \vdash \mathsf{cons}(w_{2},w_{3}) : (b \oplus c)^{*}
%     }
%   }
%   {w_{1} : a, w_{2} : b, w_{3} : c \vdash \mathsf{inl}(w_{1}) \otimes \mathsf{cons}(w_{2}, w_{3}) : (a \oplus b) \otimes (b \oplus c)^{*} }
% \end{equation}

% When parsing $w$ into $A$ by hand, we perform the same process as reflected in
% the derivation \cref{eq:samplederiv}. That is, we split the string $w$ up into pieces, one of which matches
% $a \oplus b$, and and another that matches $(b \oplus c)^{*}$. When further
% parsing this Kleene star bit, we break it up into smaller pieces each of which
% matches the patterns $b \oplus c$ and then we combine them all into one parse of
% the Kleene star $(b \oplus c)^{*}$. Note that this exact process is echoed in
% the above proof tree.

% \subsubsection{Dependence of the Linear on the Non-Linear}

% We may now however ask, ``what are the $\Gamma$'s doing in the $\otimes$ introduction rule?''
% Each linear context $\Delta$ is dependent on a non-linear context $\Gamma$, so
% we may use variables drawn from $\Gamma$ when defining the types in $\Delta$.

% In the grammar setting, this dependence lets
% us use variables from regular dependent type theory to
% keep track of data without \emph{consuming any resources}, so to speak. That is,
% using a piece of a linear context consumes it, so we cannot linearly encode anything that
% needs to be referenced more than once.

% We will see a concrete instance of this dependence in \cref{subsubsec:pda}. In
% that case, we encode the traces of a pushdown automaton as a grammar. When doing so,
% we make use of our non-linear types to define a type of stacks. The fact that
% our encoding of stacks is non-linear reflects
% that a pushdown automaton may consume a character from the input string without
% affecting the state of the stack.

% \subsubsection{Other Type Constructors}
% \label{subsubsec:other}

% \paragraph{Brzozowksi Derivative}
% In the 1950s Janusz Brzozowksi developed a notion of derivative for formal
% languages \cite{brzozowskiDerivativesRegularExpressions1964}. The derivative of language $L$ by string $w$ is given as,

% \[
%   w^{-1}L = \{ w' \in \String : w \otimes w' \in L \}
% \]

% That is, the derivative of $L$ with respect to $w$ are precisely the strings
% that complete to an $L$-parse when prefixed by a $w$.

% Brzozowski initially gave an accounting of this operation for
% generalized regular expressions, but later work demonstrates that the same
% construction can be generalized to context free grammars \cite{mightParsingDerivativesFunctional2011}.

% The ``magic wand'' from linear logic takes on the role of Brzozowski's
% derivative in our syntax.

% \begin{equation}
%   \label{eq:linfun}
% \inferrule{\Gamma ; a : A, \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A \lto B}
% \end{equation}

% We read this as saying that $\Delta$ constructs a $A \lto B$-parse when
% prepending $\Delta$ with an $A$-parse completes to a $B$ parse. In this sense,
% the type $A \lto B$ behaves like a linear function type. The only distinction
% between $A \lto B$ and an intuitionistic function type is that linear functions
% consume their input.

% Note that the introduction rule presented in \cref{eq:linfun} does indeed
% generalize Brzozowski's initial presentation of the derivative. First, to match
% his initial presentation
% let $A$ be the grammar accepting a single string $w$, which we may write as the
% discrete grammar $\lceil w \rceil$.
% Additionally, our construction lifts from a language-theoretic construction
% to a grammar point-of-view.
% In this sense, we carry around not just the data of what strings belong to
% $\lceil w \rceil \lto B$ but also are endowed the data of a function that
% constructs parse trees of $B$ from pairs of $\lceil w \rceil$-parses and
% $(\lceil w \rceil \lto B)$-parses.

% \paragraph{Top and Bottom}
% We include linear versions of the unit and empty types. The
% grammar $\top$ matches every input string, while the grammar
% $\bot$ represents falsity and matches no strings. These two
% are often used together to give syntactic shorthand to
% grammars that encode propositions.

% For instance, we may encode the proposition
% $\mathsf{isVowel}$ as,

% \[
%   \mathsf{isVowel}(w) :=
%     \begin{cases}
%       \top, \quad w \in \{a, e, i, o, u\} \\
%       \bot, \quad \text{otherwise}
%     \end{cases}
% \]

% \paragraph{Fixed Point Operator}
% We include a least fixed point operator on types, $\mu$, which is the means by
% which we define recursive grammars.

% Observe that we need not take the Kleene star as a primitive
% grammar constructor, as it is definable as a fixed point.
% The Kleene star of a grammar $g$ is given as,

% \[
%   g^{*} := \mu X . I \oplus (g \otimes X)
% \]

% \begin{figure}[h!]
% \begin{mathpar}
%   \inferrule
%   {\Gamma ; \Delta \vdash p : I}
%   {\Gamma ; \Delta \vdash \mathsf{nil} : g^{*}}

%   \and

%   \inferrule
%   {\Gamma ; \Delta \vdash p : g \\ \Gamma ; \Delta' \vdash q
%   : g^{*}}
%   {\Gamma ; \Delta \vdash \mathsf{cons}(p , q) : g^{*}}

%   \and

%   \inferrule
%   {
%     \Gamma ; \Delta \vdash p : g^{*} \\
%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
%     \Gamma ; x : g , y : h \vdash p_{\ast} : h
%   }
%   {\Gamma ; \Delta \vdash \mathsf{foldr}(p_{\varepsilon} , p_{\ast}) : g^{*}}
% \end{mathpar}
% \caption{Kleene Star Rules}
% \label{fig:star}
% \end{figure}

% Likewise, $g^{*}$ has admissible introduction and
% elimination rules, shown in \cref{fig:star}. Note that this
% definition of $g^{*}$ and these
% rules arbitrarily assigns a handedness to the Kleene star.
% We could have just as well took it to be a fixed point of
% $I \oplus (X \otimes g)$. In fact, the definitions are
% equivalent, as the existence of the $\mathsf{foldl}$ term below
% shows that $g^{*}$ is also a fixed point of
% $I \oplus (X \otimes g)$.

% \begin{equation}
%   \label{eq:foldl}
%   \inferrule
%   {
%     \Gamma ; \Delta \vdash p : g^{*} \\
%     \Gamma ; \cdot \vdash p_{\varepsilon} : h \\
%     \Gamma ; y : h , x : h \vdash p_{\ast} : h
%   }
%   {\Gamma ; \Delta \vdash \mathsf{foldl}(p_{\varepsilon} , p_{\ast}) : g^{*}}
% \end{equation}

% In fact, the $\mathsf{foldl}$ term is defined using
% $\mathsf{foldr}$ --- much in the same way one
% may define a left fold over lists in terms of a right fold
% in a functional programming language\footnote{The
%   underlying trick is to fold over a list of functions
%   instead of the original string. We curry each character $c$
%   of the string into a function that concatenates $c$, and
%   right fold over this list of linear functions. Because function
%   application is left-associative, this results in a left
%   fold over the original string. }.

% We only take fixed points of a single variable as a
% primitive operation in the type theory, but we may apply
% Beki\`c's theorem \cite{Bekić1984} to define an admissible
% notion of multivariate fixed point. This is particularly
% useful for defining grammars that encode the states of an
% automaton. In \cref{fig:multifix} we provide the
% introduction and elimination principles for such a fixed
% point, where $\sigma$ is the substitution that unrolls the
% mutually recursive definitions one level. That is,

% \begin{align*}
%   \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
%   & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
% \end{align*}

% \begin{figure}[h!]
% \begin{mathpar}
%   \inferrule
%   {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
%   {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}

%   \\

%   \inferrule
%   {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
%              \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
%              \dots \\
%              \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
%   }
%   {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
% \end{mathpar}
% \caption{Multi-fixed Points}
% \label{fig:multifix}
% \end{figure}

% The most prevalent application of these multi-fixed points
% is for defining the traces through automata, as we will see
% later in \cref{eq:nfatrace}.

% \todoin{with citation, talk somewhere about how CFGs are just regular + fixed points}
% \max{this is a good point to give an overview of how regular expressions and context-free grammars/expressions are essentially sub-languages of our type system.}

% \subsection{Semantics in Grammars}
% \label{subsec:seming}
% As mentioned earlier, the interpretation of grammars as functions from
% $\String$ to $\Set$ serves as our intended semantics. When designing the syntax of this formalism, we were often
% informed by this model. That is, there is an interpretation
% $\llbracket \cdot \rrbracket$ of our syntax that provides grammar-based
% semantics.

% \begin{enumerate}
%   \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
%   \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
%         $\sem X : \sem \Gamma \to \Set_{i}$
%   \item A non-linear term $\Gamma \vdash e : X$ denotes a section
%         $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$
%   \item Linear contexts $\Gamma \vdash \Delta$ and types
%         $\Gamma \vdash A : L_{i}$ both denote families of grammars
%         $\sem \Gamma \to \Gr_{i}$
%   \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
%         transformers
%         $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
% \end{enumerate}

% Most of the non-linear semantics is standard, so let us investigate in more
% depth the linear semantics. The interpretation of each
% connective is exactly what is suggested by the syntax, in
% large part to the inspiration taken from this interpretation
% when designing the type theory.
